---
title: "ST 551: Statistical Methods I"
date: "Lab 02 - Plots and apply()"
output: html_notebook 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(Sleuth3)
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 1 Review

Using what you learned in Lab 1, carry out the following tasks (Ask the TA for help if you get stuck):

1. Load in the `Sleuth3` library.
2. Find out what the `ex1028` data set pertains to.
3. Find the dimensions of the data set and decide how many observations and variables the data contains.
4. Which variables does R consider to be numeric (Integers in this case)? Which are considered categorical (i.e. "Factors")?
5. **Challenge**: What is the average number of hurricanes when `ElNino` is categorized as `"warm"`?


# Plots

This section focuses on building plots using the `base` graphics because it's the plotting library that ships with R when you first install it. There in an alternative graphics engine called `ggplot2` which is among the most popular R packages in existence. Eventually, I plan on showing you more `ggplot2` style graphics, but for now, we'll stick with `base` graphics since they don't require us to install any additional packages.

> You can make high-quality, publication worthy graphics with either plotting "engine". I personally prefer `ggplot2` but you should really try making plots with both to see which you prefer.

This course focuses on the most fundamental methods statisticians have developed to detect differences between, mostly two but sometimes many, populations. In my (very biased) opinion (take this with a grain of salt because it reflects the opinion of $n=1$ person), many answers to scientific questions can more easily be answered by just looking at the data visually. Plots are also a fantastic way to "explore" a data set and get a feel for the values contained within the data.

## `base` graphics

We might start out by wondering how many years in the data had an `ElNino` classification of `"cold"`, `"neutral"`, or `"warm"`. To view this visually, we can make a bargraph or barplot

```{r}
barplot(table(ex1028$ElNino))
```

### Question 1: What do you think the function `table()` does? Hint: Try just running `table(ex1028$ElNino)` to view its output.

<!-- Feel free to write your answer here -->


Previously, we saw that we can investigate the _distributions_ of quantitative variables using histograms and boxplots. So one thing we could do is look at the distribution for the number of hurricanes in our data with these plots

```{r}
hist(ex1028$Hurricanes)
boxplot(ex1028$Hurricanes)
```

### Question 2: Compare these two plots. Do you feel like they're giving you the same _view_ of the data? Or do you think one is more informative than the other? Explain your reasoning.

<!-- Feel free to write your answer here -->


### Question 3: When viewing a distribution of values, it's often important to know where the _center_ of the distribution is, it's _shape_, and also how much _spread_ there is around the center. Using one of the graphics from above, describe what you feel is the _center_, _shape_, and _spread_ for the distribution of hurricanes.

<!-- Feel free to write your answer here -->


One of the more powerful ways we can utilize plots is to visually look for relationships between variables in our data. For instance, we might wonder if the number of hurricanes in a year is related to a climate phenomenon, like El Niño.

In `base` graphics, we tend to build plots in stages, one function at a time. If we want to make three different plots for the number of hurricanes based on whether `ElNino` was classified as `"cold"`, `"neutral"`, or `"warm"` the first thing we would need to do is divide our plot panel in 3 (one for each plot). 

To divide our plot panel, we use the following:

```{r}
par(mfrow = c(1, 3))
```

The `par()` function controls many of the attributes of the plot panel. One of the arguments we can change within the `par()` function is the layout for panel as a whole. Specifically, we can change the number of plots to display in the panel (the default number of plots is one) by setting the `mfrow` argument to be some number of rows (`1` in this case) and some number of columns (`3`) in this case. The `mfrow` argument requires us to specify a _vector_ or a single R object with 1 or more elements, this is why we "combine" our number rows and columns by placing them in the `c()` function.

After we've split our single plot panel into three, the next step is to plot the three histograms. Here, we want to plot one histogram for when `ElNino` is "`warm`", one for when `ElNino` is `"cold"` and one for when `ElNino` is `"neutral"`. To accomplish this, we need to first subset our values.

We could of course subset our data using R's "bracket notation":

```{r}
warm <- ex1028[ex1028$ElNino == "warm", ]
```

The code above keeps just the rows in our `ex1028` data for which the condition `ex1028$ElNino == "warm"` is `TRUE`. This code however is rather verbose (my opinion) and also not very intuitive to "read". Another method to subset our data though is to use the `subset()` function:

```{r}
warm <- subset(ex1028, ElNino == "warm")
```

The `subset()` function wants us to specify a data set first (`ex1028` in our case), and then tell us which values we want to keep from the data set (when `ElNino == "warm"`). I would argue that this is slightly easier to read and a little bit less verbose since I don't have the type `ex1028` multiple times.

### Question 4: Run each function which subsets our `ex1028` data and verify that the subsets are the same (Consider asking a fellow student or a TA how you might verify this fact).

<!-- Feel free to write your answer here -->


Carrying on then, we can create our other subsets in a similar fashion as we used for our `warm` data.

```{r}
neutral <- subset(ex1028, ElNino == "neutral")
cold <- subset(ex1028, ElNino == "cold")
```

And with our subsetted data created, we're ready to plot! To create our three plots in one panel, all we need to do is to run the plots in order.

```{r}
hist(cold$Hurricanes)
hist(cold$Hurricanes)
hist(cold$Hurricanes)
```

> Note: In RStudio, if you run a plot command and see the error `Error in plot.new() : figure margins too large`, this means that you need to make the plot pane, as it appears in RStudio, larger.

### Question 5: Alter the plot you've just made in the following ways: (1) Have the plot display the number of hurricanes for when `ElNino` is `"cold"`, `"netural"` and `"warm"`, (2) give each plot a title denoting which temperature the plot is displaying, and (3) make the x-axis for each plot equal for each plot. To do (3), use the `xlim` argument within each `hist()` function and set it equal to `c(2, 12)`. Here, the `2` indicates where to start the x-axis and `12` denotes where it should end. 

```{r}
# Write your plot code here

```


Finally, what you might notice is that it's pretty hard to compare the distributions when they're side by side. Taking all of the code you've run so far, alter the plots so that they are displayed in `3` rows and `1` column (i.e. the histograms are on top of each other):

```{r}
# Write your plot code here

```

### Question 6: How does the category for `ElNino` impact the distribution for the number of hurricanes in a year?

<!-- Write your interpretation here (I would advise writing as good on interpretation as you can. Next year, I hope you'll remember to re-examine this interpretation to see how far you've come in interpreting statistical graphics.) -->


Finally, if you want to set the plot pane back to its default values, what I tend to do is run:

```{r}
dev.off()
```


Great work, you've (hopfully) built a really nice graphic!


# Doing things repeatedly

The last topic we cover in this lab is doing things repeatedly. The ability to do things repeatedly is one of the biggest arguments I can think of for doing work with code. Imagine the following: You're working with a data set with a large number of quantitative variables. You might want to compute the mean of each variable to get a sense of what values are _typical_ for each. If we didn't have the power to do things repeatedly, you might find yourself writing code that looks like ...

```{r}
mean(ex1028$Storms)
mean(ex1028$Hurricanes)
mean(ex1028$StormIndex)
# ... and so on ...
```

One way we can do things repeatedly is to apply a function to each column (variable) of a data set. To do this in R, there's (rather conveniently) an `apply()` function.

The `apply()` function works in the following way:

- Tell R what your data is
- Tell R whether to apply the function to the rows or columns
- Tell R which function to apply.

For example, if we wanted the `mean()` value of our `ex1028` data, we would want to tell R "On our `ex1028` data, for each column apply the `mean()` function. In practice we might run:

```{r}
apply(ex1028, 2, mean)
```

.. but as you might notice when running the code above, there's a problem. And the problem is that computing the mean of our first 4 variables isn't either (1) all the useful since the variable is just telling us the year our data set is from (Really more of a description of the data than a proper "variable") (2) Is a categorical variable (For which computing the "mean" doesn't make sense) or (3) Is a categorical variable in disguise (The variable is describing "categories" but is input as a number).

Question 7: Read the help documentation for `Temperature` and `WestAfrica` by running `?ex1028`. Why do I say that these two variables are really "categorical variables in disguise"?

<!-- Write your answer here -->

To use our apply function on just the columns where computing the `mean()` makes sense, we can either (1) Subset the data keeping just the 5th, 6th, and 7th data or (2) Subset the data by removing the first four variables.

One way we can subset the data and keep just the 5th, 6th, and 7th columns is to run:

```{r, eval = FALSE}
ex1028[, c(5, 6, 7)]
```

> Notice: Here, notice that you've just been introduced to a new way to subset the data! You can combine the row numbers, or column numbers, you want to keep within a `c()` function and use R's bracket notation to keep just those rows or columns! Fancy.

If we want to remove the first four columns of our data, we can run
```{r, eval = FALSE}
ex1028[, -c(1, 2, 3, 4)]
```

One more tip: Writing a series of integers is a pain. In R, we can write `c(1, 2, 3, 4)` as `1:4` which gives us an equivalent output (Don't believe me, run `c(1,2, 3, 4)` and `1:4` in the console!)

### Question 8: Take the `apply()` function we ran above but this time subset the data so that we compute the `mean()` for just columns 5, 6, and 7.

```{r}
# Alter the code here
apply(ex1028, 2, mean)
```


One last bit about the `apply()` function. If you wanted to apply a function to each row, we would change the 2nd argument (currently `2`) to `1`. In R, rows are typically considered the first dimension (hence why we would use `1` in the apply function) while columns are considered the second dimension (hence the `2` in our current `apply()` code).